---
title: "Project1"
author: "Damon D'Ambrosio"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE)
library(tidyverse)
library(jsonlite)
library(httr)
```

## Requirements

Packages required to run the functions are as follows:  
- `tidyverse`: All manner of good things for manipulating and analyzing data  
- `httr`: Connecting to the API  
- `jsonlite`: Getting usable information from the API

## API Interaction Custom Functions

### `pokemon.lookup`

The first custom function we have is `pokemon.lookup`. This function receives an input, either in the form of a Pokemon’s name or ID number, and returns a nice, neat, one-row tibble displaying the desired information for that individual Pokemon. PokeApi does not require a key, so queries are accomplished by simply pasting the input query to the end of the necessary hyperlink to look up the information. Name queries on PokeApi are case-sensitive and lowercase only, so some language is provided in the function to allow the user to enter the pokemon name however they see fit. By default, PokeApi stores Pokemon height and weight in decimeters and hectograms, respectively. I have a hard enough time conceptualizing metric units, and these particular units do not match those used in any of the Pokemon-related media I remember from my youth. Similarly, they do not match the units used in Pokemon-related media made for consumption outside of the United States. As such, an optional `unit` argument exists, which will convert the height and weight into imperial units (inches and pounds) or metric units (meters and kilograms). `pokemon.lookup` is a nice standalone function, but will also serve as a helper function to generate larger reports. 

```{r lookup_function}
pokemon.lookup <- function(id, unit = NULL){
  poke.query <- paste0("https://pokeapi.co/api/v2/pokemon/", str_to_lower(id))
  
  poke.get <- GET(poke.query)
  
  poke.char <- fromJSON(rawToChar(poke.get$content))
  
  name <- str_to_title(poke.char$name)
  id.number <- poke.char$id
  type1 <- poke.char$types$type$name[1]
  type2 <- poke.char$types$type$name[2]
  height <- poke.char$height
  weight <- poke.char$weight
  hp <- poke.char$stats$base_stat[1]
  attack <- poke.char$stats$base_stat[2]
  defense <- poke.char$stats$base_stat[3]
  special.attack <- poke.char$stats$base_stat[4]
  special.defense <- poke.char$stats$base_stat[5]
  speed <- poke.char$stats$base_stat[6]
  
  if("imperial" %in% unit){
    height <- round((height * 3.93700787))
    weight <- round((weight * 0.2204622622))
  }
  
  if("metric" %in% unit){
    height <- round((height/10), 1)
    weight <- round((weight/10), 1)
  }

  poke.table <- tibble(name = name,
                       id.number = id.number,
                       type1 = type1,
                       type2 = type2,
                       height = height,
                       weight = weight,
                       hp = hp,
                       attack = attack,
                       special.attack = special.attack,
                       special.defense = special.defense,
                       speed = speed
                )
return(poke.table)
}
```

Lets test it out. My favorite Pokemon as a kid was Rhydon, so let’s make a query for Rhydon, using imperial units:

```{r lookup_example1}
pokemon.lookup("rhydon", unit = "imperial")
```

Now lets demonstrate how it functions with using a numeric Pokemon ID and metric units. How about one of the 300 first pokemon, selected randomly?
```{r lookup_example2}
pokemon.lookup(sample(1:300,1), unit = "metric")
```

## Data Exploration

A narrative showing a data exploration activity will go here. I have a preliminary idea of what I would like to do, depending on the information available in the API.
