---
title: "Project-1"
author: "Damon D'Ambrosio"
date: "`r Sys.Date()`"
output:
  github_document:
    html_preview: false
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE)
library(tidyverse)
library(jsonlite)
library(httr)
```

## Requirements

Packages required to run the functions are as follows:  
- `tidyverse`: All manner of good things for manipulating and analyzing data  
- `httr`: Connecting to the API  
- `jsonlite`: Getting usable information from the API

## API Interaction Custom Functions

### `pokemon.lookup`

The first custom function we have is `pokemon.lookup`. This function receives an input, either in the form of a Pokemon’s name or ID number, and returns a nice, neat, one-row tibble displaying the desired information for that individual Pokemon. PokeApi does not require a key, so queries are accomplished by simply pasting the input query to the end of the necessary hyperlink to look up the information. Name queries on PokeApi are case-sensitive and lowercase only, so some language is provided in the function to allow the user to enter the Pokemon name however they see fit. By default, PokeApi stores Pokemon height and weight in decimeters and hectograms, respectively. I have a hard enough time conceptualizing metric units, and these particular units do not match those used in any of the Pokemon-related media I remember from my youth. Similarly, they do not match the units used in Pokemon-related media made for consumption outside of the United States. As such, an optional `unit` argument exists, which will convert the height and weight into imperial units (inches and pounds) or metric units (meters and kilograms). `pokemon.lookup` is a nice standalone function, but will also serve as a helper function to generate larger reports. 

```{r lookup_function}
pokemon.lookup <- function(x, unit = NULL, ...){
  poke.query <- paste0("https://pokeapi.co/api/v2/pokemon/", str_to_lower(x))
  
  poke.get <- GET(poke.query)
  
  poke.char <- fromJSON(rawToChar(poke.get$content))
  
  name <- str_to_title(poke.char$name)
  id.number <- poke.char$id
  type1 <- poke.char$types$type$name[1]
  type2 <- poke.char$types$type$name[2]
  height <- poke.char$height
  weight <- poke.char$weight
  hp <- poke.char$stats$base_stat[1]
  attack <- poke.char$stats$base_stat[2]
  defense <- poke.char$stats$base_stat[3]
  special.attack <- poke.char$stats$base_stat[4]
  special.defense <- poke.char$stats$base_stat[5]
  speed <- poke.char$stats$base_stat[6]
  
  if("imperial" %in% unit){
    height <- round((height * 3.93700787))
    weight <- round((weight * 0.2204622622))
  }
  
  if("metric" %in% unit){
    height <- round((height/10), 1)
    weight <- round((weight/10), 1)
  }

  poke.table <- tibble(name = name,
                       id.number = id.number,
                       type1 = type1,
                       type2 = type2,
                       height = height,
                       weight = weight,
                       hp = hp,
                       attack = attack,
                       defense = defense,
                       special.attack = special.attack,
                       special.defense = special.defense,
                       speed = speed
                )
return(poke.table)
}
```

Lets test it out. My favorite Pokemon as a kid was Rhydon, so let’s make a query for Rhydon, using imperial units:

```{r lookup_example1}
pokemon.lookup("rhydon", unit = "imperial")
```

Now lets demonstrate how it functions with using a numeric Pokemon ID and metric units. How about one of the 300 first Pokemon, selected randomly?

```{r lookup_example2}
pokemon.lookup(sample(1:300,1), unit = "metric")
```

The ultimate plan is to use `lapply` on `pokemon.lookup` to generate large reports for analysis. However, there is some functionality I would like to add. Namely, I do not have Pokemon ID numbers memorized. I do remember Pokemon names, however. As such, it would be nice if I could simply request all Pokemon between "Pokemon A" and "Pokemon B" using names, and have the function handle looking up the ID, constructing the vector, then feeding it to `lapply`. The next custom function deals with that.

###  `pokemon.vector`

`pokemon.vector` allows one to provide the names of two Pokemon (via a character vector with length = 2), and returns a vector of the ID numbers between said Pokemon. This vector can then be fed to `pokemon.lookup` via `lapply` to generate a report of multiple Pokemon.

```{r vector_function}
pokemon.vector <- function(x, ...){
  
  if(is.numeric(x)){
    stop("Expecting two Pokemon names, not numbers!")
  }
  
  if(length(x) > 2){
    stop("Expecting exactly two Pokemon names!")
  }
  
  poke.1 <- str_to_lower(x[1])
  poke.2 <- str_to_lower(x[2])
  
  poke.1.query <- paste0("https://pokeapi.co/api/v2/pokemon/", poke.1)
  poke.1.get <- GET(poke.1.query)
  poke.1.char <- fromJSON(rawToChar(poke.1.get$content))
  poke.1.id <- poke.1.char$id
  
  poke.2.query <- paste0("https://pokeapi.co/api/v2/pokemon/", poke.2)
  poke.2.get <- GET(poke.2.query)
  poke.2.char <- fromJSON(rawToChar(poke.2.get$content))
  poke.2.id <- poke.2.char$id
  
  poke.vec <- seq(poke.1.id, poke.2.id)
  return(poke.vec)
}
```

Now, a quick demonstration of `pokemon.vector` in action. Let's get a numeric vector of all Pokemon IDs of the Generation III Pokemon. "Treecko" is the first Generation III Pokemon, and "Jirachi" is the last Generation III Pokemon:

```{r vector_example}
gen3.list <- c("Treecko", "Jirachi")
pokemon.vector(gen3.list)
```

One last helper function before we wrap things up...

### `gen1.correction`

The Pokemon games have changed a lot from since I was a kid. Some of these changes were to correct "balance" issues. The two major changes that come to mind are as follows:

- New Pokemon "types" have been added since Generation I to correct imbalances in the strengths/weaknesses of the various Pokemon. "Steel" and "Dark" types were added in Generation II, and at some point in time, "Fairy" was also added as a type. Seven Generation I Pokemon retroactively had their types changed as a result. I would like to do an exploratory data analysis on the Pokemon as they existed in Generation I, so would need to correct these Pokemon to their "old" typing.

- As can be seen earlier in this vignette, Pokemon have a "special attack" and "special defense" stat. In Generation I, these were combined into a single stat called "special." From what I can tell, the original "Special" stat does not exist in PokeApi. However, it can exist within R as a simple numeric vector with 151 values. The `gen1.correction` function will store and provide that information.

```{r gen1_correct}
gen1.correction <- function(x, type.change = TRUE, add.special = TRUE, delete.sp.att.def = FALSE, ...){

y <- x %>% filter(id.number <= 151)
z <- x %>% filter(id.number > 151)

  if(isTRUE(type.change)){
    y$type1[y$type1 == "fairy"] <- "normal" #Corrects Clefairy and Clefable to Gen I typing
    y$type2[y$type2 == "fairy"] <- NA       #Corrects Jigglypuff, Wigglytuff, and Mr-Mime to Gen I typing
    y$type2[y$type2 == "steel"] <- NA       #Corrects Magnemite and Magneton to Gen I typing
    
    if(max(x$id.number) > 151){
      warning("Pokemon that did not exist in Generation I are in this report. The typing scheme in this report is now inconsistent")
    }
  }
  
  if(isTRUE(add.special)){
  
  #Special stat values sourced from bulbapedia.bulbagarden.net
  special.vector <- c(65, 80, 100, 50, 65, 85, 50, 65, 85, 20, 25, 80, 20, 25, 45, 35, 50, 70, 25, 50, 31, 61, 40, 65, 50, 90, 30, 55, 40, 55,   75, 40, 55, 75, 60, 85, 65, 100, 25, 50, 40, 75, 75, 85, 100, 55, 80, 40, 90, 45, 70, 40, 65, 50, 80, 35, 60, 50, 80, 40, 50, 70, 105, 120,   135, 35, 50, 65, 70, 85, 100, 100, 120, 30, 45, 55, 65, 80, 40, 80, 95, 120, 58, 35, 60, 70, 95, 40, 65, 45, 85, 100, 115, 130, 30, 90, 115,   25, 50, 55, 80, 60, 125, 40, 50, 35, 35, 60, 60, 85, 30, 45, 105, 100, 40, 70, 95, 50, 80, 70, 100, 100, 55, 95, 85, 85, 55, 70, 20, 100,     95, 48, 65, 110, 110, 110, 75, 90, 115, 45, 70, 60, 65, 125, 125, 125, 50, 70, 100, 154, 100)

  y <- y %>% mutate(gen1.special = special.vector[y$id.number])
  }

new.report <- bind_rows(y,z)

  if(isTRUE(delete.sp.att.def)){
    if(max(new.report$id.number > 151)){
      warning("Pokemon that did not exist in Generation I are in this report. Are you sure you wanted to delete special attack and special defense?")
    }
  new.report <- new.report %>% select(-c(special.attack, special.defense))
  }
return(new.report)
}

```

`gen1.correction` will check that Pokemon belong to Generation I by making sure their ID number is 151 or lower. It will then change types if needed, and append the "Special" stat to that Pokemon's record. If a type change is requested and Pokemon that did not exist in Generation I are included in the provided data set, a warning message will notify the user that the overall type scheme is inconsistent. The user can also delete "Special Attack" and "Special Defense" if they wish to have a "pure" Generation I dataset. If Pokemon that did not exist in Generation I is included in the provided data set and the user specifies dropping "Special Attack" and "Special Defense," a warning message will appear, as the resultant data set will have no "special" data associated with such records.

Here's a quick example. Let's build a simple data set containing two Pokemon: Magnemite, a Generation I Pokemon who would need a type correction to remove its secondary "Steel" typing, and Scizor, a Generation II Pokemon who does not need a type correction, but would be inadvertently "corrected" if the function did not work as intended. We'll also specify the function to delete Special Attack and Special Defense to see both potential warning messages:

```{r gen1_correction_example, warning = TRUE}
a <- pokemon.lookup("Magnemite")
b <- pokemon.lookup("Scizor")
c <- bind_rows(a,b)

gen1.correction(c, delete.sp.att.def = TRUE)
```

As we can see, Magnemite had its secondary Steel type assigned to NA, while Scizor retained its secondary Steel type. A warning message was printed since a Pokemon outside of Generation I was in the report while Special Attack and Special Defense were deleted. Finally, Scizor does not have a `gen1.special` value.

### `pokemon.batch.report`

The final step is to put all of these functions into a nice wrapper, generate some data, and start an exploratory analysis.

```{r batch_lookup}
pokemon.batch.report <- function(x, gen1.correct = FALSE, ...){
  if(!is.numeric(x)){
    if(length(x) > 1){
    x <- pokemon.vector(x)
    }
  }
  
  x <- lapply(x, pokemon.lookup, ...)
  
  batch.lookup <- bind_rows(x)
  
  if(isTRUE(gen1.correct)){
    batch.lookup <- gen1.correction(batch.lookup, ... )
  }
  
  return(batch.lookup)
  
}

```

Now to use `pokemon.batch.report` to generate some data for exploratory analysis. As a millennial approaching middle age, the Generation I Pokemon and the games they appeared in are nearest and dearest to my heart. As such, I want the original 151 Pokemon (Bulbasaur through Mew). While one *could* supply `pokemon.batch.report` with a vector like `x <- 1:151` to generate these data, I will supply the two Pokemon names to take advantage of the `pokemon.vector` helper function. I will request imperial units, and have the data "corrected" to reflect Generation I typing. For part of the exploratory analysis, I would like to compare how the change from a "special" stat to a separate "special attack" and "special defense" impacted the overall abilities of each Pokemon. As such, I will retain "special attack" and "special defense" when applying the Generation I correction.

```{r batch_report_demo}
#To test the full capabilities of the wrapper, we will provide it with a vector of two Pokemon names instead of ID numbers
my.data <- pokemon.batch.report(c("Bulbasaur","Mew"), gen1.correct = TRUE, unit = "imperial")

print(my.data)
```

## Data Exploration

### Pokemon type frequency breakdown

Lets start with a contingency table. One of the criticisms of the Generation I games was the distribution of Pokemon typing; some types had many Pokemon, whereas others had very little. Pokemon always have at least one type, and can have up to two types. Let's see how that breakdown actually looks:

```{r type_table}
#Including useNA option to return Pokemon with no secondary typing
type.table <- table(my.data$type1, my.data$type2, useNA = "ifany")

print(type.table)
```

The table shows this claim has some merit. There are no "pure" (i.e, `type2` = `NA`) Ghost-, Ice-, or Rock-type Pokemon in Generation I. Conversely, there are many "pure" Water-, Normal-, Poison-, and Fire-type Pokemon. Let's generate some percent totals to better explore type frequency. First, lets get the row sums of the above table, and calculate the percent of the 151 total Generation I Pokemon that had a particular type as their primary type (`type1`):

```{r type1_frequency}
type1.table <- type.table %>%
  as.data.frame() %>%
    as_tibble %>%
      rename(type1 = Var1, type2 = Var2) %>%
        group_by(type1) %>%
          summarize(total = sum(Freq)) %>%
            mutate(pct.total = round(((total/151)*100),1)) %>%
              arrange(desc(pct.total))

print(type1.table)
```
As we can see, a whopping 28 Generation I Pokemon (~18.5%) had Water as their primary type. Conversely, there were only 2 Pokemon with Ice as their primary type (~1.3%). Let's repeat the process for the secondary type (`type2`):
```{r type2_table}
type2.table <- type.table %>%
  as.data.frame() %>%
    as_tibble %>%
      rename(type1 = Var1, type2 = Var2) %>%
        group_by(type2) %>%
          summarize(total = sum(Freq)) %>%
            mutate(pct.total = round(((total/151)*100),1)) %>%
              arrange(desc(pct.total))

print(type2.table)
```
As we can see, the majority of Generation I Pokemon do not have a secondary type (i.e., `type2` = `NA`). Flying and Poison are tied for the most common secondary type. Flying was not a type that appeared in our type1 table, and Normal was not a type that appeared in our type2 table. This indicates some types in Generation I only exist as primary types, whereas others only exist as secondary types. Let's see which types fall into these categories. First, types that only appear as a primary type (`type1`):

```{r type1_only}
#Rename columns in preparation for setdiff
type1.table <- rename(type1.table, type = type1)
type2.table <- rename(type2.table, type = type2)

#Extract columns
type1.column <- type1.table$type
type2.column <- type2.table$type

#Compare
setdiff(type1.column, type2.column)
```
No Generation I Pokemon had a secondary typing of Normal, Bug, Fire, Electric, Dragon, or Ghost. Let's look at `type2`:

```{r type2_only}
#Compare
setdiff(type2.column, type1.column)
```
No Generation I Pokemon had a non-existent primary typing (as expected, as every Pokemon has a primary type). Further, no Pokemon has Flying as its primary type.

### Height and Weight

While some attacks introduced in later Pokemon games were affected by Pokemon height and weight, in Generation I, this was not the case. As such, heights and weights are really more of general interest. As a kid, I remember there being an assumption that Rock- and Ground-type Pokemon were generally heavier, and some of the Flying-type Pokemon were lighter. Let's see if there's any truth to that:

```{r weight_summary}
my.data %>% group_by(type1) %>% 
  summarize(average = mean(weight), median = median(weight), min = min(weight), max = max(weight), sd = sd(weight)) %>% 
  arrange(desc(average))

```

It appears Rock-type Pokemon are indeed the heaviest on average, with an average weight of 193 pounds. Ground-type Pokemon, however, are well within the middle of the pack. Based on comparisons of the mean and median, it looks like Generation I Rock-type Pokemon weights are right-skewed, whereas the corresponding distribution for Ground-type Pokemon is left skewed. There appears to be a 2 pound Ground-type Pokemon who is probably responsible for some of that. Let's see who it is:

```{r light_ground}
my.data %>% filter(type1 == "ground") %>% filter(weight == 2) %>% select(1:6)
```

Ah, yes, Diglett. I remember that being a small Pokemon.

As indicated earlier, Flying only existed as a secondary type in Generation 1. Let's do another summary to see how the data for Pokemon with a secondary Flying type shakes out:

```{r flying_summary}
my.data %>% group_by(type2) %>% 
  summarize(average = mean(weight), median = median(weight), min = min(weight), max = max(weight), sd = sd(weight)) %>%
  arrange(desc(average))
```
Interesting. Flying-type Pokemon shake out in the middle of the pack. Apparently there are some very heavy flying things in the Pokemon universe, and a bit of right-skew in the distribution. Out of curiosity, let's see them all together:

```{r heavy_flying}
my.data %>% filter(type2 == "flying") %>% arrange(desc(weight)) %>% select(1:6)
```

Ah, I forgot Gyarados had a secondary Flying type. Even discounting Gyarados, and the two other Pokemon here whose designs are not inspired by birds/insects/bats (i.e., Dragonite and Charizard), there are still some pretty heavy flying things in the Generation I universe.

Let's take a quick look at height. I really have no idea what, if any, type I would expect to be taller:

```{r height_summary}
my.data %>% group_by(type1) %>% 
  summarize(average = mean(height), median = median(height), min = min(height), max = max(height), sd = sd(height)) %>% 
  arrange(desc(average))
```

Not as much parity in heights across type compared to what was seen with weights. If I remember correctly from my youth, there are a few serpent-looking Pokemon here that are probably skewing data, as their "length" is deemed their "height" in the Pokemon universe. Let's take a quick look at all Pokemon with a "height" greater than 7 feet:

```{r height_list}
my.data %>% filter(height > 84) %>% arrange(desc(height)) %>% select(1:6)
```

As predicted. Onix, Gyarados, Dragonair, and Arbok, the four "tallest" Pokemon, are all serpent-like, and "height" would probably more accurately be described as "length." Anyway, enough of the height and weight business; let's get to the interesting data...

### Base Stats

The remainder of the data returned by the `pokemon.batch.report` function are the "base stats" of each Pokemon species. Briefly, the whole purpose of the Pokemon games is to capture Pokemon, train them, and then use them to battle other Pokemon. Each of these base stats gives an idea of how proficient a Pokemon is at a particular aspect that can be used in battle. Although these stats will increase as the Pokemon is trained, the base stats give a general idea of how one Pokemon will compare to another (i.e., a Pokemon with a higher base speed stat will be faster than another with a lower base speed stat, all other things considered). A quick overview of each stat:

- **HP** is how many "hit points" a Pokemon has; hit points are depleted when attacked, so a Pokemon with a higher hp stat can withstand more damage.
- **Attack** is how much damage a Pokemon will dole out when attacking, assuming the attack is of a certain type
- **Defense** is how much a Pokemon can reduce the damage it receives when attacked, assuming the attack is of a certain type
- **Speed** is how fast a Pokemon is; the Pokemon games are turn-based, and faster Pokemon usually attack first
- **Special/Special Attack/Special Defense** will be discussed a little later in the vignette

Let's use some histograms to get a better idea of how distributed these stats are among the 151 Generation I Pokemon. First, Attack:

```{r attack_histogram}
g <- ggplot(my.data, aes(attack))

g + geom_histogram(binwidth = 10, color = "black", fill = "royalblue1", size = 1) + xlab("Base Attack") + theme_classic()

```

Attack looks like it has a fairly normal distribution, maybe with a hint of bimodality. Let's check it by primary Pokemon type with a boxplot:

```{r attack_boxplot}
#Create an array of colors that roughly correspond to the colors associated with types in the games
poke.colors <- c('olivedrab3', 'lightblue4', 'yellow', 'sienna', 'red', 'slateblue4', 'green3', 'tan', 'slategray1', 'ivory2', 'mediumpurple1', 'mediumorchid2', 'wheat4', 'steelblue2')

g <- ggplot(my.data, aes(x=type1, y = attack, fill = type1))

g + geom_boxplot() + 
  xlab("Type") + ylab("Base Attack") + 
  scale_x_discrete(guide = guide_axis(angle = 45)) + 
  theme_classic() + 
  scale_fill_manual(values = poke.colors) +
  theme(legend.position = "none")
```

It appears that within Generation I, there are no "outliers" within each type as it relates to the base Attack stat. Fighting-type Pokemon stand out as having overall high attacks, while Ghost- and Psychic-type Pokemon stand out as having lower median Attack values. There are pretty large ranges of Attack within the Bug- Normal- and Water-types. Overall though, attack across type seems relatively balanced.

Let's look at defense. First, the overall histogram:

```{r}
g <- ggplot(my.data, aes(defense))

g + geom_histogram(binwidth = 10, color = "black", fill = "royalblue1", size = 1) + xlab("Base Defense") + ylab("Count") + theme_classic()
```

Looks like some right skew, especially when compared to the Attack graph. Let's check the boxplot:

```{r}
g <- ggplot(my.data, aes(x=type1, y = defense, fill = type1))

g + geom_boxplot() + xlab("Type") + ylab("Base Defense") + 
  scale_x_discrete(guide = guide_axis(angle = 45)) + 
  theme_classic() + 
  scale_fill_manual(values = poke.colors) + 
  theme(legend.position = "none")
```

Here, we see Rock-type Pokemon, and Ground-type to a lesser extent, stand out as those with the highest Defense stat. We do see potential outliers here, namely a Water-type Pokemon with a very high Defense.

Let's check speed the same way. Histogram first:

```{r}
g <- ggplot(my.data, aes(speed))

g + geom_histogram(binwidth = 10, color = "black", fill = "royalblue1", size = 1) + xlab("Base Speed") + theme_classic()
```

Somewhat normal, with perhaps a touch of right skew.

Now boxplots:

```{r}
g <- ggplot(my.data, aes(x=type1, y = speed, fill = type1))

g + geom_boxplot() + 
  xlab("Type") + 
  ylab("Base Speed") +
  scale_x_discrete(guide = guide_axis(angle = 45)) + 
  theme_classic() + 
  scale_fill_manual(values = poke.colors) + 
  theme(legend.position = "none")
```

It appears Electric- and Psychic-type Pokemon are fast, and there is one really slow Water-type Pokemon.

Now for HP. Histogram:

```{r}
g <- ggplot(my.data, aes(hp))

g + geom_histogram(binwidth = 10, color = "black", fill = "royalblue1", size = 1) + xlab("Base HP") + theme_classic()
```

Lots of skew. There appear to be a few Pokemon with a ton of base hp. Now the boxplots:

```{r}
g <- ggplot(my.data, aes(x=type1, y = hp, fill = type1))

g + geom_boxplot() + xlab("Type") + ylab("Base HP") + 
  scale_x_discrete(guide = guide_axis(angle = 45)) + 
  scale_fill_manual(values = poke.colors) + 
  theme_classic() + 
  theme(legend.position = "none")
```

Both of those high HP Pokemon appear to be Normal-type. Now, let's wrap up with a brief look and discussion of the "Special" stats.

#### Special Stats

As discussed at points throughout this vignette, the API provides a "Special Attack" and "Special Defense" stat for each Pokemon. Special Attack and Special Defense are similar to Attack and Defense, with one slight difference. In addition to all Pokemon having a type, each "move" they can learn also has a type. The potential move types are the same as the potential Pokemon types. In Generation I (and Generations II and III; I believe this has changed as the franchise continues to expand), Water-, Grass-, Fire-, Ice-, Electric-, and Psychic-type moves are considered "special" type moves. As an example, the amount of damage a Pokemon can inflict with a Water-type move would be determined by its Special Attack, not its Attack. Similarly, the ability of its opponent to withstand the damage caused by the move would be determined by the opponent's Special Defense, not its Defense.

In Generation I, however, separate stats for Special Attack and Special Defense did not exist. They were combined into a single stat that was simply called "Special." This caused issues with "balance" in the games, as a Pokemon with a high Special could both dole out heavy damage if using a special-type move, and similarly, do an admirable job of withstanding special type attacks. Splitting this stat into Special Attack and Special Defense helped correct the issue.

I am curious to see how each Generation I Pokemon was affected by this "split" of the Special stat. My idea for investigating this is as follows. Consider the following scatter plot, which shows Special Attack versus Special Defense (with a y=x line added for reference):

```{r split_special_scatter}
g <- ggplot(my.data, aes(x=special.attack, y = special.defense, color = type1))

g + geom_abline(slope = 1, intercept = 0) + 
  geom_point() + 
  xlab("Special Attack") + ylab("Special Defense") + 
  scale_color_manual(name = "Type", values = poke.colors) + 
  theme_classic()
```

Assuming Special Attack and Special Defense are equally useful, a Pokemon can thought to be "stronger" in terms of Special stats the further away it is from the hypothetical (0,0) point. The figure above shows a nice split in the number of Pokemon "above" the reference line (i.e., better Special Defense) and "below" the reference line (i.e., better Special Attack). This is how the figure would look if considering the Generation I Special stat, wherein "Special Attack" and "Special Defense" are identical:

```{r gen1_special_scatter}
g <- ggplot(my.data, aes(x=gen1.special, y = gen1.special, color = type1))

g + geom_abline(slope = 1, intercept = 0) + 
  geom_point() + 
  xlab("Gen 1 'Special Attack'") + ylab("Gen 1 'Special Defense'") + 
  scale_color_manual(name = "Type", values = poke.colors) + 
  theme_classic()
```

As expected, a not particularly illuminating visual of a straight y=x line. With some application of the Pythagorean theorem, we can calculate the "distance" each Pokemon resides from the hypothetical (0,0) under both the modern assignment of Pokemon stats, with separate Special Attack/Special Defense, and under the Generation I singular Special stat. In other words, after Generation I:


$new.distance = \sqrt{special.attack^2 + special.defense^2}$

And within Generation I:

$old.distance = \sqrt{2(special^2)}$

Let's derive these variables, calculate the difference, and see how Pokemon were "improved" or "weakened" by the splitting the Special stat.

```{r distance_variables}
my.data <- my.data %>% mutate(new.distance = sqrt((special.attack^2) + (special.defense^2)),
                              old.distance = sqrt(2*(gen1.special^2)),
                              delta.distance = new.distance - old.distance)
```

Under these calculations, a Pokemon with a positive `delta.distance` was "improved" when the Special stat was split, a Pokemon with a negative `delta.distance` was "weakened," and a Pokemon with 0 `delta.distance` was unchanged. Let's add in another variable that assigns discrete levels to reflect this:

```{r delta_distance_status_variable}
my.data <- my.data %>% mutate(status = ifelse(delta.distance > 0, "improved",
                                              ifelse(delta.distance <0, "weakened", "neutral")))
```

Let's take a quick look at some contingency tables. First, an overall view of how many Pokemon belong to each `status`:

```{r delta_distance_table}
table(my.data$status)

```

It looks like roughly 1/3 of the original Pokemon were "improved," 1/3 were "weakened," and the remaining 1/3 were unchanged. Let's see how that balance holds up across type:

```{r delta_distance_type_table}
table(my.data$type1, my.data$status)
```

It does not hold up. It looks like some types, such as Psychic, were overall weakened by this stat change. Others, like Normal, appear to have improvement. All Ghost-type Pokemon were weakened, whereas almost all Fighting-type Pokemon were improved. Let's see a bar graph to better understand the actual amount of change:

```{r delta_distance_barplot}
g <- ggplot(my.data, aes(x = type1, y = delta.distance, fill = type1))

g + geom_hline(yintercept = 0) + 
  geom_bar(stat = "summary", fun = "mean", color = "black") + 
  xlab("Type") + ylab("Mean Change after Special Stat Split") + 
  scale_x_discrete(guide = guide_axis(angle = 45)) + 
  scale_fill_manual(values = poke.colors) + 
  theme_classic() + 
  theme(legend.position = "none")
```

From this graph, it appears Ghost- and Psychic-type Pokemon were pretty weakened by the stat split. Conversely, Fighting-type Pokemon saw great improvement. Dragon- and Ground-type Pokemon appear to be unaffected by the split. I'm curious as to which specific Pokemon were most affected:

```{r delta_distance_list}
my.data %>% arrange(delta.distance) %>% select(1:4, 16)
```

Ah, Mewtwo. He was (and still is, as far as I know) the "strongest" Generation I Pokemon, so he could afford to be "weakened" a bit. Here's how each Generation I Pokemon changed as a result of the Special split:

```{r}
g <- ggplot(my.data, aes(x = name, y = delta.distance, fill = type1))

g + geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = position_dodge(.2)) + 
  ylab("Change after Special Stat Split") + 
  scale_x_discrete(guide = guide_axis(angle = 90)) + 
  scale_fill_manual(name = "Type", values = poke.colors) + 
  theme_classic() + 
  theme(legend.position = "top" , axis.text.x = element_text(size = 3), axis.title.x = element_blank())
```

The graph shows Mewtwo being "weakened" the most of all. Two Fighting-type Pokemon, Hitmonchan and Hitmonlee, appeared to benefit greatly from the Special split. Overall, as the initial contingency table showed, there is a pretty nice division of "weakened" and "improved" Pokemon.

## Conclusion
Thus concludes the vignette. While it was interesting to look at some of the height, weight, and type distribution stuff, I had an honest curiosity about the base stats, especially as it relates to "splitting" the Special stat in later generations, and the subsequent effect on the overall "strength" of each Pokemon. I found it interesting to explore, and hopefully you found it somewhat interesting to read.